Hybrid Partial Least Squares Regression with Multiple Functional and Scalar Predictors

### 2025-08-13 manuscript/main.tex
score orthogonal proposition revised

	A fundamental property of partial least squares   is that its derived bases are orthonormal. Our regularized estimates preserve this property, with respect to a modified inner product that incorporates the roughness penalty, defines as follows:
	\begin{definition}[Roughness-sensitive inner product]\label{def:hybrid_inner_product_roughness}
		Given two hybrid predictors $W_1 = (X_{11}, \ldots, X_{1K}, \mathbf{Z}_1)$ and $W_2 = (X_{21}, \ldots, X_{2K}, \mathbf{Z}_2)$, both elements of $\mathcal{H}$ as defined in Definition \ref{def:hybid_predictor}, and a roughness penalty matrix $\Lambda = \operatorname{blkdiag}(\lambda_1 I_M, \ldots, \lambda_K I_M, 0_{p \times p})$, the roughness-sensitive inner product between $W_1$ and $W_2$ is defined as:
		\begin{equation}		\label{eq: hybrid inner product_roughness}
			\langle W_1, W_2\rangle_{\mathcal{H}, \Lambda}
			:=
			\sum \limits_{k=1}^K \int_0^1 X_{1k}(t) X_{2k}(t) \, dt
			+
			\sum \limits_{k=1}^K \lambda_k \int_0^1 X^{\prime \prime}_{1k}(t) X^{\prime \prime}_{2k}(t) \, dt
			+
			\mathbf{Z}_1^\top \mathbf{Z}_2.
		\end{equation}
	\end{definition}
	Based on this inner product, the following proposition states that the PLS component directions estimated from Proposition  \ref{proposition:eigen_regul} are orthonormal.
	\begin{proposition}[Orthonormality of regularized PLS component directions]\label{proposition: modified orthnormality of PLS components}
		The PLS component directions $\widehat{\xi}_1, \widehat{\xi}_2, \ldots, \widehat{\xi}_L$, estimated via Proposition \ref{proposition:eigen_regul} with a roughness penalty matrix $\Lambda = \operatorname{blkdiag}(\lambda_1 I_M, \ldots, \lambda_K I_M, 0_{p \times p})$, are mutually orthonormal with respect to the inner product $\langle \cdot, \cdot \rangle_{\mathcal{H}, \Lambda}$. That is,
		\begin{equation*}
			\langle \widehat{\xi}_{l_1}, \widehat{\xi}_{l_2} \rangle_{\mathcal{H}, \Lambda}
			= \mathbbm{1}(l_1 = l_2), \quad l_1, l_2 = 1, \ldots, L.
		\end{equation*}
	\end{proposition}
	The proof of Proposition \ref{proposition: modified orthnormality of PLS components} is provided in Appendix \ref{section:proof:proposition: modified orthnormality of PLS components}.

	\jongmin{revising here}
	Now let $\widehat{\boldsymbol{\rho}}_l=(\widehat{\rho}_{1l}, \ldots, \widehat{\rho}_{nl})^\top$ denote the $n$-dimensional vector whose elements consist of estimated $l$-th PLS scores ($l=1, \ldots, L)$ of $n$ observations. 
	The next proposition states that the vectors of estimated PLS scores $\widehat{\boldsymbol{\rho}}_1, \widehat{\boldsymbol{\rho}}_2, \ldots, \widehat{\boldsymbol{\rho}}_L$ obtained from the proposed algorithm are mutually orthogonal.
	\begin{proposition}	\label{proposition: orthnormality of PLS scores}
		The vectors of estimated PLS scores, $\widehat{\boldsymbol{\rho}}_1, \widehat{\boldsymbol{\rho}}_2, \ldots, \widehat{\boldsymbol{\rho}}_L$, obtained from the proposed PLS algorithm are mutually orthogonal in the sense that
		\begin{equation*}
			\widehat{\boldsymbol{\rho}}_l^\top \widehat{\boldsymbol{\rho}}_j= 0 \quad \textnormal{for} \quad \; l,j \in \{1, \ldots, L\}, \; l \ne j. 
		\end{equation*}
	\end{proposition}
	The proof of Proposition \ref{proposition: orthnormality of PLS scores} is provided in Appendix \ref{section:proof:proposition: orthnormality of PLS scores}.

@@

		Our next task is to show that 

@@

@@

	\bigr)^\top \in \tilde{\mathcal{H}}^{\otimes n}.

@@

Then using \eqref{algorithm_step:residualization_predictor:recall}, we have, with some abuse of notation,

@@

Clearly, the right-hand side represents a linear operator from $\tilde{\mathcal{H}}^{\otimes n}$ to $\tilde{\mathcal{H}}^{\otimes n}$, which we denote as $P^{[l_1-1]}$.

@@

	we have $\widetilde{\boldsymbol{W}}^{[l_1]}  = 
	P^{[l_1-1]} \widetilde{\boldsymbol{W}}^{[l_1-1]}$.
Using this relationship, we have

@@

		\widetilde{\boldsymbol{W}}^{[l_1]} &=P^{[l_1-1]} \widetilde{\boldsymbol{W}}^{[l_1-1]} \\

@@

		P^{[l_1-1]} P^{[l_1-2]} \widetilde{\boldsymbol{W}}^{[l_1-2]}

@@

		&= P^{[l_1-1]} P^{[l_1-2]} \ldots P^{[l_2 + 1]}  P^{[l_2]} \widetilde{\boldsymbol{W}}^{[l_2]}

@@

&=P P^{[l_2]} \widetilde{\boldsymbol{W}}^{[l_2]},		

@@

	where $P = P^{[l_1-1]} P^{[l_1-2]} \ldots P^{[l_2 + 1]} $ is a linear operator from  $\tilde{\mathcal{H}}^{\otimes n}$.
	Then for $j < l$,
		\int_{\mathcal{T}} \widetilde{W}^{[l]}[\mathbf{t}] \widehat{\xi}_j[\mathbf{t}] d\mathbf{t} &= \int_{\mathcal{T}} P   \left(I_n - \frac{\widehat{\boldsymbol{\rho}}_{j} \widehat{\boldsymbol{\rho}}_{j}^\top}{\widehat{\boldsymbol{\rho}}_{j}^\top \widehat{\boldsymbol{\rho}}_{j}} \right)\widetilde{W}^{[j]}[\mathbf{t}]  \widehat{\xi}_k[\mathbf{t}] d\mathbf{t}  \\
		&= P \left\{\int_{\mathcal{T}} \widetilde{W}^{[j]}[\mathbf{t}] \widehat{\xi}_j[\mathbf{t}] d\mathbf{t} - \frac{\widehat{\boldsymbol{\rho}}_{j} \widehat{\boldsymbol{\rho}}_{j}^\top}{\widehat{\boldsymbol{\rho}}_{j}^\top \widehat{\boldsymbol{\rho}}_{j}} \int_{\mathcal{T}} \widetilde{W}^{[j]}[\mathbf{t}] \widehat{\xi}_j[\mathbf{t}] d\mathbf{t} \right\} \\ 
		&=P \left( \widehat{\boldsymbol{\rho}}_j - \frac{\widehat{\boldsymbol{\rho}}_{j} \widehat{\boldsymbol{\rho}}_{j}^\top}{\widehat{\boldsymbol{\rho}}_{j}^\top \widehat{\boldsymbol{\rho}}_{j}} \widehat{\boldsymbol{\rho}}_j \right) \\
		&= P(\widehat{\boldsymbol{\rho}}_j-\widehat{\boldsymbol{\rho}}_j) \\
		&= \mathbf{0} \in \mathbb{R}^n, 
	noting that $\int_{\mathcal{T}} \widetilde{W}^{[j]}[\mathbf{t}] \widehat{\xi}_j[\mathbf{t}] d\mathbf{t} = \widehat{\boldsymbol{\rho}}_j$. Now by the , we have that

	\begin{align*}
		&=  \frac{1}{\kappa_l}\widehat{\xi}_l^{* \top} V^\ast_l \widehat{\xi}_j^{*} \\
		&= \frac{1}{\kappa_l}\widehat{\xi}_l^{* \top} \left(n^{-2} J^\ast \widetilde{C}^{*[l]\top} \widetilde{\mathbf{y}}^{[l]}\widetilde{\mathbf{y}}^{[l]\top} \widetilde{C}^{*[l]} J^\ast \right) \widehat{\xi}_j^{*} \\
		&= \frac{1}{n^2\kappa_l}\widehat{\xi}_l^{* \top} \ J^\ast \widetilde{C}^{*[l]\top} \widetilde{\mathbf{y}}^{[l]}\widetilde{\mathbf{y}}^{[l]\top}  \widetilde{C}^{*[l]}\left(\int_{\mathcal{T}} B^*[\mathbf{t}] B^*[\mathbf{t}]^\top d\mathbf{t}  \right) \widehat{\xi}_j^{*} \\
		&= \frac{1}{n^2\kappa_l}\widehat{\xi}_l^{* \top}  J^\ast \widetilde{C}^{*[l]\top} \widetilde{\mathbf{y}}^{[l]}\widetilde{\mathbf{y}}^{[l]\top}  \left(\int_{\mathcal{T}} \widetilde{C}^{*[l]} B^*[\mathbf{t}] B^*[\mathbf{t}]^\top \widehat{\xi}_j^{*} d\mathbf{t}  \right)   \\
		&= \frac{1}{n^2\kappa_l}\widehat{\xi}_l^{* \top}  J^\ast \widetilde{C}^{*[l]\top} \widetilde{\mathbf{y}}^{[l]}\widetilde{\mathbf{y}}^{[l]\top}  \left(\int_{\mathcal{T}} \widetilde{W}^{[l]}[\mathbf{t}] \widehat{\xi}_j[\mathbf{t}] d\mathbf{t}  \right)   \\
		&=0, 
	\end{align*}
	where the last equality follows from the above result that $\int_{\mathcal{T}} \widetilde{W}^{[l]}[\mathbf{t}] \widehat{\xi}_j[\mathbf{t}] d\mathbf{t} = \mathbf{0}$.

	\end{proof}
	\section{Proof of Proposition \ref{proposition: orthnormality of PLS scores}}\label{section:proof:proposition: orthnormality of PLS scores}

>>>>

	A fundamental property of partial least squares   is that its derived bases are orthonormal. Our regularized estimates preserve this property, with respect to a modified inner product that incorporates the roughness penalty, defines as follows:
	\begin{definition}[Roughness-sensitive inner product]\label{def:hybrid_inner_product_roughness}
		Given two hybrid predictors $W_1 = (X_{11}, \ldots, X_{1K}, \mathbf{Z}_1)$ and $W_2 = (X_{21}, \ldots, X_{2K}, \mathbf{Z}_2)$, both elements of $\mathcal{H}$ as defined in Definition \ref{def:hybid_predictor}, and a roughness penalty matrix $\Lambda = \operatorname{blkdiag}(\lambda_1 I_M, \ldots, \lambda_K I_M, 0_{p \times p})$, the roughness-sensitive inner product between $W_1$ and $W_2$ is defined as:
		\begin{equation}		\label{eq: hybrid inner product_roughness}
			\langle W_1, W_2\rangle_{\mathcal{H}, \Lambda}
			:=
			\sum \limits_{k=1}^K \int_0^1 X_{1k}(t) X_{2k}(t) \, dt
			+
			\sum \limits_{k=1}^K \lambda_k \int_0^1 X^{\prime \prime}_{1k}(t) X^{\prime \prime}_{2k}(t) \, dt
			+
			\mathbf{Z}_1^\top \mathbf{Z}_2.
		\end{equation}
	\end{definition}
	Based on this inner product, the following proposition states that the PLS component directions estimated from Proposition  \ref{proposition:eigen_regul} are orthonormal.
	\begin{proposition}[Orthonormality of estimated PLS component directions]\label{proposition: modified orthnormality of PLS components}
		The PLS component directions $\widehat{\xi}_1, \widehat{\xi}_2, \ldots, \widehat{\xi}_L$, estimated via Proposition \ref{proposition:eigen_regul} with a roughness penalty matrix $\Lambda = \operatorname{blkdiag}(\lambda_1 I_M, \ldots, \lambda_K I_M, 0_{p \times p})$, are mutually orthonormal with respect to the inner product $\langle \cdot, \cdot \rangle_{\mathcal{H}, \Lambda}$. That is,
		\begin{equation*}
			\langle \widehat{\xi}_{l_1}, \widehat{\xi}_{l_2} \rangle_{\mathcal{H}, \Lambda}
			= \mathbbm{1}(l_1 = l_2), \quad l_1, l_2 = 1, \ldots, L.
		\end{equation*}
	\end{proposition}
	The proof of Proposition \ref{proposition: modified orthnormality of PLS components} is provided in Appendix \ref{section:proof:proposition: modified orthnormality of PLS components}.

The next proposition states that the vectors of estimated PLS scores for different iteration numbers are mutually orthogonal.
	\begin{proposition}	\label{proposition: orthnormality of PLS scores}
		Recall from Lemma \ref{proposition:closed_form_orthgonalization} that   $\widehat{\boldsymbol{\rho}}^{[l]}$ denote the $n$-dimensional vector whose elements consist of the $l$-th estimated PLS scores ($l=1, \ldots, L$) of $n$ observations.
		The vectors $\widehat{\boldsymbol{\rho}}^{[1]}, \widehat{\boldsymbol{\rho}}^{[2]}, \ldots, \widehat{\boldsymbol{\rho}}^{[L]}$  are mutually orthogonal in the sense that
		\begin{equation*}
			\widehat{\boldsymbol{\rho}}^{[l_1] \top}  \widehat{\boldsymbol{\rho}}^{[l_2] \top}= 0 \quad \textnormal{for} \quad \; l_1,l_2 \in \{1, \ldots, L\}, \; l_1 \ne l_2. 
		\end{equation*}
	\end{proposition}
	The proof of Proposition \ref{proposition: orthnormality of PLS scores} is provided in Appendix \ref{section:proof:proposition: orthnormality of PLS scores}.

@@

Orthogonality can be demonstrated by showing that $\Theta^{[l_1] } \mathbb{B} \hat{\boldsymbol{\xi}}_{l_2} = \mathbf{0} \in \mathbb{R}^n$.

@@

			\numberthis \label{thetaBzetal2i}

@@

	\bigr)^\top \in \tilde{\mathcal{H}}^{\otimes n},

@@

and
	\langle 	\widetilde{\boldsymbol{W}}^{[l_1]}  ,
	\rangle_{\mathcal{H}}
	\widetilde{W}_1^{[l_1-1]}, 
	\widetilde{W}_n^{[l_1-1]},  
	\rangle_{\mathcal{H}}
	\biggr)^\top 
	\in \mathbb{R}^n.
Recalling \eqref{thetaBzetal2i} our goal of showing $
is equivalent to showing  $	\langle 	\widetilde{\boldsymbol{W}}^{[l_1]}  ,
\rangle_{\mathcal{H}} = \mathbf{0}$.
Using this notation and  \eqref{algorithm_step:residualization_predictor:recall}, we have, with some abuse of notation,

@@

and thus
\langle 	\widetilde{\boldsymbol{W}}^{[l_1]}  ,
\rangle_{\mathcal{H}}
\langle 	\widetilde{\boldsymbol{W}}^{[l_1-1]}  ,
\rangle_{\mathcal{H}} 
- 
	\frac{1
	}{
	\| 	\hat{\boldsymbol{\rho}}^{[l_1-1]}\|_2^2
		\hat{\boldsymbol{\rho}}^{[l_1-1]}
	\hat{\boldsymbol{\rho}}^{[l_1-1] \top}
	\bigl \langle 
	\widetilde{\boldsymbol{W}}^{[l_1-1]}  ,
	\bigr \rangle_{\mathcal{H}}.
Clearly, the right-hand side represents a linear operator from $\mathbb{R}^n$ to $\mathbb{R}^n$, which we denote as $P^{[l_1-1]}$.

@@

	we have $\langle 	\widetilde{\boldsymbol{W}}^{[l_1]}  ,
	\rangle_{\mathcal{H}}  = 
	P^{[l_1-1]} \langle 	\widetilde{\boldsymbol{W}}^{[l_1-1]}  ,
	\rangle_{\mathcal{H}}$.
Repeatedly using this relationship, we have

@@

	\langle 	\widetilde{\boldsymbol{W}}^{[l_1]}  ,
	\rangle_{\mathcal{H}} 

@@

P^{[l_1-1]} \langle 	\widetilde{\boldsymbol{W}}^{[l_1-1]}  ,
\rangle_{\mathcal{H}} 

@@

		P^{[l_1-1]} P^{[l_1-2]}   \langle 	\widetilde{\boldsymbol{W}}^{[l_1-2]}  ,
		\rangle_{\mathcal{H}}

@@

		&= \underbrace{ P^{[l_1-1]} P^{[l_1-2]} \ldots P^{[l_2 + 1]}}_{:=P}  P^{[l_2]} \langle 	\widetilde{\boldsymbol{W}}^{[l_2]}  ,
		\rangle_{\mathcal{H}}
&= P \,\biggl\{
\langle 	\widetilde{\boldsymbol{W}}^{[l_2]}  ,
\rangle_{\mathcal{H}} 
- 
\frac{1
}{
	\| 	\hat{\boldsymbol{\rho}}^{[l_2]}\|_2^2
\hat{\boldsymbol{\rho}}^{[l_2]}
\hat{\boldsymbol{\rho}}^{[l_2] \top}
\bigl \langle 
\widetilde{\boldsymbol{W}}^{[l_2]}  ,
\bigr \rangle_{\mathcal{H}}
&= P \,\biggl\{
\hat{\boldsymbol{\rho}}^{[l_2]}
- 
\frac{1
}{
	\| 	\hat{\boldsymbol{\rho}}^{[l_1-1]}\|_2^2
\hat{\boldsymbol{\rho}}^{[l_2]}
\hat{\boldsymbol{\rho}}^{[l_2] \top}
 \hat{\boldsymbol{\rho}}^{[l_2]}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\\&= P \,\biggl\{
\hat{\boldsymbol{\rho}}^{[l_2]}
- 
\frac{\hat{\boldsymbol{\rho}}^{[l_2] \top} \hat{\boldsymbol{\rho}}^{[l_2]}
}{
	\| 	\hat{\boldsymbol{\rho}}^{[l_1-1]}\|_2^2
\hat{\boldsymbol{\rho}}^{[l_2]}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\\&= P \,\biggl\{
\hat{\boldsymbol{\rho}}^{[l_2]}
- 
\hat{\boldsymbol{\rho}}^{[l_2]}
P \mathbf{0} = \mathbf{0}.

@@

This completes the proof of Proposition 
	\ref{proposition: modified orthnormality of PLS components}.
	\section{Proof of Proposition \ref{proposition: orthnormality of PLS scores}}\label{section:proof:proposition: orthnormality of PLS scores}
